## How the numbers work

Each tab below walks through one step of the analysis on a few rows of toy data: the
input, the code that turns it into a score, the output, and how to read it. The code calls
the same functions the analysis scripts use (from the project's `R/` folder), so the
arithmetic is the real thing; only the data are made up and small enough to check by hand.

```{r we-setup}
#| include: false
# Worked examples: load the pipeline's own functions (no data needed).
for (f in c("features.R", "cluster.R", "evaluate.R", "gsi.R", "gsi_model.R", "cutb_curves.R"))
  sys.source(file.path("..", "R", f), envir = globalenv())
options(digits = 3, width = 100)
we_ok <- exists("r2_groups") && exists("gm_score") && exists("whit_fit")
```

::: {.panel-tabset}

### 1. Per-year features

**What it is for.** Every climate feature is computed for each year first and only then
summarized across years, as the median (typical value) and the interquartile range (IQR,
year-to-year spread). Averaging the climate first would give the wrong answer for anything
nonlinear, such as a threshold or a ramp.

**Input.** One grid cell, eight years of last spring freeze dates (day of year) and
water-year precipitation.

```{r we-1a}
#| echo: true
yr  <- data.frame(year = 2018:2025,
                  last_freeze_doy = c(98, 112, 105, 87, 120, 101, 95, 109),
                  p_ann_mm = c(310, 190, 420, 260, 150, 390, 280, 230))
yr
# summary across years, as in summaries/block1_median.tif and block1_iqr.tif
sapply(yr[-1], function(v) c(median = median(v), IQR = IQR(v)))
```

**Why not average the climate first?** A GSI-style ramp is 0 below a threshold and 1 above
it. Here is a 28-day precipitation ramp (10 to 20 mm, the FEMS setting) applied to three
years, then to their average:

```{r we-1b}
#| echo: true
p28 <- c(dry = 4, normal = 12, wet = 35)          # 28-day precipitation (mm) in three years
ramp <- m_ramp(p28, 10, 20)                       # the pipeline's ramp function
ramp
c(mean_of_ramp = mean(ramp), ramp_of_mean = m_ramp(mean(p28), 10, 20))
```

The average year (17 mm) scores 0.70, but the three years average 0.40: one year
switched the index off completely. That is why the GSI is run every year and then
summarized.

**Aridity index.** Precipitation divided by Hargreaves potential evapotranspiration (PET),
which needs only daily minimum and maximum temperature and the latitude:

```{r we-1c}
#| echo: true
ra <- ra_matrix(lat = 31.8, doy = 182)            # top-of-atmosphere radiation, July 1, SE Arizona
pet <- hargreaves(tmin = 21, tmax = 37, ra = ra)  # mm/day
c(Ra_MJ = ra, PET_mm_per_day = pet)
# a water year with 330 mm of rain and ~1,700 mm of PET
c(aridity_index = 330 / 1700)
```

Values below 0.65 are drylands; most of the West is well below that, so the index is
log-transformed before clustering.

### 2. Rain onset

**What it is for.** Block 2 records when the warm-season rains start, because monsoon
grasses green up after the first soaking rains, not on a calendar date. The rule: the first
wet day (at least 1 mm) from May 1 that starts a 10-day total of at least 20 mm, with no dry
spell of 20 days or more in the following 30 days (a false start).

**Input.** Daily rain for one toy summer: a false start in mid June, then the real onset in
early July.

```{r we-2a}
#| echo: true
doy <- 121:273                                     # May 1 .. Sep 30
ppt <- setNames(rep(0, length(doy)), doy)
ppt[as.character(c(165, 166, 168))] <- c(12, 9, 6) # mid-June storms (27 mm in 4 days) ...
                                                   # ... then 25 dry days
ppt[as.character(c(194, 195, 197, 199, 203, 206, 210, 214, 218))] <- c(8, 14, 5, 11, 7, 9, 12, 6, 10)
onset <- function(ppt, doy, wet = 1, window = 10, total = 20, check = 30, max_dry = 20) {
  for (i in seq_along(ppt)) {
    if (ppt[i] < wet) next
    if (sum(ppt[i:min(length(ppt), i + window - 1)]) < total) next
    nxt <- ppt[(i + 1):min(length(ppt), i + check)] < wet          # dry days in the next 30
    runs <- rle(nxt); if (any(runs$values & runs$lengths >= max_dry)) next   # false start
    return(doy[i])
  }
  max(doy) + 1                                     # no onset: censored at Oct 1
}
onset(ppt, doy)
```

Day 165 (June 14) starts 27 mm in ten days but is followed by a 25-day dry spell, so it is
rejected. Day 194 (July 13) is the onset. Years with no onset are recorded as October 1,
and the share of years with an onset is a feature of its own.

### 3. The GSI

**What it is for.** The Growing Season Index multiplies daily ramps (0 = fully limiting,
1 = not limiting) for minimum temperature, vapor pressure deficit (VPD), daylength and,
optionally, moisture, then averages the product over 21 or 28 days.

**One day, FEMS settings.** Minimum temperature −2 to 5 °C, VPD 3,882 down to 1,956 Pa,
daylength 11 to 12 h, 28-day precipitation 10.16 to 20.32 mm.

```{r we-3a}
#| echo: true
day <- c(tmin = 3.5, vpd_pa = 2600, daylength_h = 13.2, p28_mm = 14)
sub <- c(tmin  = m_ramp(day[["tmin"]], -2, 5),
         vpd   = m_ramp(day[["vpd_pa"]], 1956, 3882, decreasing = TRUE),
         photo = m_ramp(day[["daylength_h"]], 11, 12),
         moist = m_ramp(day[["p28_mm"]], 10.16, 20.32))
round(sub, 2)
round(prod(sub), 2)                               # the day's index (iGSI)
names(which.min(sub))                             # limiting factor
```

The day's index is 0.20, and moisture is the limiting factor (the smallest ramp). Block 3's
*limiting factor* features count the share of days each ramp is the smallest one below 1.

**Season dates from a smoothed index.** A toy year: dormant in winter, a green season in
summer, a dry spell in between. The dates are read the same way as for the satellite curves:
green-up when the index rises past 20 % of the season's range, curing when it falls below
50 % (relative). The NFDRS4 reading uses fixed levels instead: green-up at the threshold GU
and half cured at (1 + GU) / 2.

```{r we-3b}
#| echo: true
d <- 1:365
igsi <- pmax(0, sin((d - 120) / 160 * pi)) * ifelse(d %in% 200:215, 0.3, 1)   # daily iGSI
G <- matrix(m_roll_mean(matrix(igsi, 1), 28), 1)                              # 28-day mean
Gweak <- G * 0.5                                                              # a poor year
g <- list(season_halfwidth_days = 183, min_amp = 0.05)
rbind(relative_20_50      = gm_season_dates(G, d, d, g, "rel")[1, ],
      nfdrs4_gu0.3        = gm_season_dates(G, d, d, g, "abs_nfdrs4", c(gu = 0.3))[1, ],
      nfdrs4_gu0.6        = gm_season_dates(G, d, d, g, "abs_nfdrs4", c(gu = 0.6))[1, ],
      weak_relative_20_50 = gm_season_dates(Gweak, d, d, g, "rel")[1, ],
      weak_nfdrs4_gu0.3   = gm_season_dates(Gweak, d, d, g, "abs_nfdrs4", c(gu = 0.3))[1, ])
```

Two things to notice. With fixed levels, the mid-summer dry spell already counts as curing
(day 213 with GU 0.3), while the relative reading waits for the real decline (day 269). And
in the poor year the index never reaches the NFDRS4 curing level, so there is no curing
date at all, while the relative reading still gives one. That difference is behind the
scoring issue in the curing test (tab 10).

**KBDI.** The Keetch–Byram Drought Index is a running soil-moisture deficit (0 = saturated,
800 = maximum drought). It climbs with heat and drops with rain, so it remembers dry spells
longer than a 28-day rain sum.

```{r we-3c}
#| echo: true
days <- 120
P  <- matrix(c(rep(0, 60), 25, rep(0, 29), 40, rep(0, 29)), 1)   # mm/day: two storms
TX <- matrix(rep(35, days), 1)                                     # hot: 35 C every day
k <- m_kbdi(P, TX, R_in = 12)                                      # mean annual rain 12 in
round(k[1, c(1, 30, 60, 61, 90, 91, 120)])          # days 1, 30, 60, 61, 90, 91, 120
```

The index climbs from 10 to about 415 over 60 hot, dry days; the 25 mm storm on day 61
takes off about 70, the 40 mm storm on day 91 about 130, and it climbs again after each.

### 4. Satellite curves

**What it is for.** Cut B describes each cell's season from MODIS. Composites with clouds
or snow read too low, so the smoother is fitted, the points well below the curve are given
less weight, and it is fitted again (an *upper envelope*). Season dates are then read from
the smooth curve.

**Input.** A toy year of NDVI every 8 days with two cloudy composites.

```{r we-4a}
#| echo: true
set.seed(4)
node <- seq(1, 361, by = 8)
true <- 0.25 + 0.35 * exp(-((node - 200) / 45)^2)                   # green peak near July 19
obs <- true + rnorm(length(node), 0, 0.015)
obs[c(20, 27)] <- obs[c(20, 27)] - 0.20                             # two cloudy composites
fit <- whit_envelope(matrix(obs, 1), matrix(1, 1, length(node)), lambda = 30, iter = 1, below_weight = 0.2)
round(rbind(observed = obs, smoothed = fit[1, ])[, 17:30], 2)       # May to August
```

```{r we-4b}
#| echo: true
z <- fit; pk <- which.max(z); base <- min(z); amp <- max(z) - base
m4 <- c(peak_doy = node[pk], base = base, amplitude = amp,
  sos20 = cc_crossing(z, pk, base + 0.2 * amp, node, 30, TRUE),    # green-up: 20 % up the rise
  eos50 = cc_crossing(z, pk, base + 0.5 * amp, node, 30, FALSE))   # end: 50 % down the fall
round(m4, 2)
```

The two cloudy composites (`r sprintf("%.2f and %.2f", obs[20], obs[27])`) barely pull the
smooth curve down (`r sprintf("%.2f and %.2f", fit[1, 20], fit[1, 27])`). Green-up (sos20) is
where the curve passes 20 % of the way from its base to its peak, here about day
`r round(m4[["sos20"]])`; the end of season (eos50) is where it falls back below 50 %, about
day `r round(m4[["eos50"]])`.
Curing (cure50) is read the same way from the NDII7 curve, at 50 % of its decline after its
peak.

### 5. Feature groups

**What it is for.** Before clustering, every feature is standardized, then reduced by
principal components within its group, and each group is scaled to the same total variance.
Otherwise the group with the most features dominates: Block 1 has eight temperature
features but only three for moisture amount.

**Input.** Six cells: eight nearly identical temperature features and three moisture
features (cells 1–3 are wet, 4–6 dry).

```{r we-5a}
#| echo: true
set.seed(5)
t0 <- c(8, 9, 10, 8, 9, 10)                       # temperature signal
m0 <- c(2.9, 3.0, 2.8, 2.3, 2.2, 2.4)             # log10 precipitation signal
X <- cbind(sapply(1:8, function(i) t0 + rnorm(6, 0, 0.1)), sapply(1:3, function(i) m0 + rnorm(6, 0, 0.02)))
colnames(X) <- c(paste0("temp", 1:8), paste0("moist", 1:3))
one  <- groups_reduce(X, NULL, 1, 0.99)                                   # one group
two  <- groups_reduce(X, list(temperature = paste0("temp", 1:8), moisture = paste0("moist", 1:3)), 1, 0.99)
share <- function(r) { v <- colSums(r$scores^2); round(v / sum(v), 2) }
share(one); share(two)
```

As one group, the first component is mostly temperature and moisture comes second. In two
groups, each gets half of the variance, so the clustering sees the wet/dry split as clearly
as the warm/cool one. Blocks are weighted the same way (Block 1 and Block 2 equal; elevation
and latitude at a quarter).

### 6. Stability

**What it is for.** A set of regions is only useful if it does not change much when the
data change. Stability fits k-means twice on two different random halves of the map (drawn
as 2-degree blocks), labels the same evaluation cells with both, and compares the two
labellings with the **adjusted Rand index** (ARI): 1 = identical regions, about 0 = no more
alike than chance.

```{r we-6a}
#| echo: true
a <- c(1, 1, 1, 2, 2, 2, 3, 3, 3)
c(same_regions_renumbered = ari(a, c(3, 3, 3, 1, 1, 1, 2, 2, 2)),
  one_cell_moved          = ari(a, c(1, 1, 2, 2, 2, 2, 3, 3, 3)),
  unrelated               = ari(a, c(1, 2, 3, 1, 2, 3, 1, 2, 3)))
```

Numbering does not matter, only who is grouped with whom. Moving one of nine cells already
drops the index to 0.64, and an unrelated grouping can go below zero. A2's stability of 0.83
at 7 regions means two fits on different halves of the country agree more closely than that. The same index compares two different sets of regions (A3 vs A2:
0.71–0.80).

### 7. Curve variance

**What it is for.** How much of the variation in the satellite seasonal curves lies between
regions rather than within them. It is an R-squared: 1 − (within-region sum of squares) /
(total sum of squares), summed over all curve features (the principal-component scores of
the NDVI and NDII7 curves).

**Input.** Six cells with two curve features, and two ways of grouping them.

```{r we-7a}
#| echo: true
Y <- cbind(pc1 = c(2.0, 1.8, 2.2, -1.9, -2.1, -2.0), pc2 = c(0.3, -0.2, 0.1, 0.4, -0.1, -0.5))
climate_regions  <- c(1, 1, 1, 2, 2, 2)       # follows the curves
location_regions <- c(1, 1, 2, 2, 3, 3)       # compact blobs that cut across them
c(climate = r2_groups(Y, climate_regions), location = r2_groups(Y, location_regions))
```

The *location only* baseline is the second kind: compact regions of the same number with no
climate information. A set of climate regions earns its keep by beating it, which A1, A2 and
A3 do on timing (about 0.55 against 0.47 at 13 regions in herbaceous cells).

### 8. Ground observations

**What it is for.** Whether observed dates at NPN, live fuel moisture and iNaturalist sites
differ more between regions than within them. Each site is first reduced to its median date
across years. The score is the **adjusted R-squared** of those dates on region, which
subtracts what any random grouping into that many regions would explain by chance:
adj = 1 − (1 − R²)(n − 1)/(n − m), with n sites and m regions.

```{r we-8a}
#| echo: true
sites <- data.frame(site = 1:12,
                    green_doy = c(95, 102, 99, 110, 160, 171, 158, 166, 120, 131, 125, 140),
                    region = c(1, 1, 1, 1, 2, 2, 2, 2, 3, 3, 3, 3),
                    lon = c(-104, -104, -103, -102, -111, -111, -110, -110, -97, -97, -96, -96),
                    lat = c(40, 41, 41, 40, 32, 32, 33, 33, 39, 39, 38, 36))
r2_adj(sites$green_doy, sites$region)
r2_adj(sites$green_doy, rep(1:6, 2))                     # 6 arbitrary groups
```

The real regions explain 95 % of the spread (adjusted 0.93). Six arbitrary groups still
explain 32 % by chance, and the adjustment takes that away (−0.24).

**How sure is a difference?** Nearby sites are not independent, so sites are resampled in
2-degree blocks (whole blocks in or out), 500 times, and both sets of regions are scored on
each resample. The table on the page reports the difference with the range covering 90 % of
resamples; a range that spans zero means the observations cannot tell the two apart.

```{r we-8b}
#| echo: true
set.seed(8)
blk <- paste(floor(sites$lon / 2), floor(sites$lat / 2)); idx <- split(seq_len(nrow(sites)), blk)
other <- c(1, 1, 3, 3, 2, 2, 2, 2, 3, 3, 3, 1)           # a second, worse set of regions
d <- replicate(500, { pick <- unlist(idx[sample(names(idx), length(idx), TRUE)])
  r2_adj(sites$green_doy[pick], sites$region[pick])[["adj_r2"]] - r2_adj(sites$green_doy[pick], other[pick])[["adj_r2"]] })
b8 <- c(difference = r2_adj(sites$green_doy, sites$region)[["adj_r2"]] - r2_adj(sites$green_doy, other)[["adj_r2"]],
        quantile(d, c(0.05, 0.95), na.rm = TRUE), share_above_0 = mean(d > 0, na.rm = TRUE))
round(b8, 2)
```

The first set of regions scores `r sprintf("%.2f", b8[["difference"]])` higher, and
`r sprintf("%.0f %%", 100 * b8[["share_above_0"]])` of resamples favour it, but the range
(`r sprintf("%+.2f", b8[[2]])` to `r sprintf("%+.2f", b8[[3]])`)
`r if (b8[[2]] < 0) "spans zero: with twelve sites in a handful of blocks the difference is not established" else "stays above zero"`. That is the situation for most of the grass comparisons on the page.

### 9. Synchrony

**What it is for.** Whether a region's cells green up and cure early or late *in the same
years*. Each cell's yearly date becomes a departure from its own median; the score is the
share of all departures that the region's mean departure in the same year explains.
Departures over 90 days are dropped as season jumps (for example a monsoon cell whose peak
switches to spring).

**Input.** Departures (days) for four cells over five years, in two regions. The function is
copied from the synchrony script.

```{r we-9a}
#| echo: true
sync_r2 <- function(a, g) {                       # as in scripts/33_synchrony.R
  ok <- !is.na(g); a <- a[ok, , drop = FALSE]; g <- g[ok]
  tot <- sum(a^2, na.rm = TRUE); res <- 0
  for (j in seq_len(ncol(a))) {                   # each year: departures from the region mean
    x <- a[, j]; u <- !is.na(x); mu <- tapply(x[u], g[u], mean)
    res <- res + sum((x[u] - mu[as.character(g[u])])^2)
  }
  1 - res / tot
}
A <- rbind(c(-10, 5, 12, -3, -4),    # region 1: cells move together
           c(-8, 7, 10, -6, -3),
           c(6, -9, 2, 11, -10),     # region 2: cells move independently
           c(-7, 4, -8, -2, 13))
c(two_regions = sync_r2(A, c(1, 1, 2, 2)), one_conus_region = sync_r2(A, c(1, 1, 1, 1)))
```

Region 1 is almost perfectly in step, region 2 not at all, and the total lands in between.
Across CONUS at 13 regions the real values are about 0.25 for green-up and 0.15 for curing.

### 10. GSI model test

**What it is for.** Whether a GSI fitted to a region predicts the satellite green-up and
curing dates in years it has not seen. Thresholds are fitted on odd years and scored on even
years, then the reverse. Scores:

- **MAE**, mean absolute error in days, each error capped at 90 days (season jumps);
- **within 30 days**, the share of years predicted within a month;
- **anomaly correlation**, whether the model's early and late years match the observed ones
  (each cell's dates minus its own mean);
- **coverage**, the share of observed dates for which the GSI gives a date at all;
- the **typical date** (the cell's median date in the fitting years) as the baseline to beat.

**Input.** Observed and predicted green-up and curing dates for three cells over five
held-out years. The GSI tracks the first two cells; in the third (think monsoon Southwest)
it gives a curing date in only two years.

```{r we-10a}
#| echo: true
obs <- list(sos  = rbind(c(100, 110, 95, 120, 105), c(150, 160, 170, 140, 155), c(160, 175, 150, 185, 165)),
            cure = rbind(c(200, 215, 190, 230, 205), c(250, 262, 240, 270, 255), c(180, 240, 175, 235, 185)))
gsi <- list(sos  = rbind(c(104, 112, 100, 115, 101), c(155, 150, 168, 150, 158), c(170, 168, 160, 175, 171)),
            eos  = rbind(c(205, 220, 196, 228, 200), c(255, 250, 247, 265, 258), c(NA, 232, NA, 241, NA)))
typical <- list(sos = rbind(rep(106, 5), rep(155, 5), rep(166, 5)),       # median of the fitting years
                eos = rbind(rep(209, 5), rep(255, 5), rep(200, 5)))
sc <- rbind(gsi_penalty  = gm_score(gsi, obs, penalty = 60, cap = 90),                   # first curing test
            gsi_fallback = gm_score(gsi, obs, penalty = 60, cap = 90, fill = typical),   # rerun
            typical_date = gm_score(typical, obs, penalty = 60, cap = 90, anom = FALSE))
round(sc[, c("sos_mae", "sos_anom_r", "cure_mae", "cure_within30", "cure_coverage",
             "cure_anom_r", "cure_paired_mae", "cure_paired_fill_mae")], 2)
```

```{r we-10-text}
#| include: false
r10 <- round(sc, 1)
```

Reading the curing columns:

- With the **fixed penalty** (first curing test) each of the three missing years counts as a
  60-day miss, and the GSI scores `r r10["gsi_penalty", "cure_mae"]` days against
  `r r10["typical_date", "cure_mae"]` for the typical date. In a cell where any guess would
  miss by more than 60 days, a fit can score better by giving no date at all: that is what
  happened in the monsoon Southwest.
- With the **fallback** (the rerun) a missing year takes the cell's typical date, so
  skipping can never beat the baseline: the GSI scores `r r10["gsi_fallback", "cure_mae"]`
  days. **Coverage** (`r r10["gsi_fallback", "cure_coverage"]`) shows how often the GSI dated
  the year itself, and the **paired** columns compare GSI and typical date only on the years
  the GSI dated: `r r10["gsi_fallback", "cure_paired_mae"]` against
  `r r10["gsi_fallback", "cure_paired_fill_mae"]` days. That paired comparison is the
  honest test of skill.
- The **anomaly correlation** uses only the years the GSI dated, and asks whether its early
  and late years line up with the observed ones. The typical date has none by definition:
  it predicts the same date every year.

**Fitting.** The thresholds are found by minimizing the same capped error over the fitting
years with a simplex search (Nelder–Mead), started from the FEMS values. A one-parameter
version: find the precipitation threshold whose green-up dates best match the observed ones.

```{r we-10b}
#| echo: true
p28 <- c(0, 0, 1, 3, 6, 9, 12, 15, 18, 22, 25, 28)        # 28-day rain, 12 ten-day steps
obs_day <- 70                                              # observed green-up
loss <- function(th) { first <- which(p28 >= th)[1]; if (is.na(first)) 60 else abs(first * 10 - obs_day) }
sapply(c(5, 10, 13, 20), loss)                             # error (days) for four thresholds
optimize(loss, c(0, 30))$minimum                           # a best threshold
```

Any threshold between 9 and 12 mm puts green-up on day 70. The error changes in steps (a
date moves a whole day or not at all), which is why a search that compares points
(Nelder–Mead) is used rather than one that follows slopes. With five to nine thresholds at
once, the search has to stop somewhere; the first curing test stopped at 250 tries per
search in almost every fit, the rerun allows 600.

### 11. Profiles

**What it is for.** The profile tables summarize each region by the median and IQR of its
cells, and the crosswalk shows how fine regions sit inside coarse ones (the share of each
fine region's cells in each coarse region).

```{r we-11a}
#| echo: true
fine   <- c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3)
coarse <- c(1, 1, 1, 2, 2, 2, 2, 2, 2, 1)
tab <- table(fine = fine, coarse = coarse)
round(prop.table(tab, 1), 2)                     # rows: share of each fine region's cells
p_ann <- c(310, 280, 350, 330, 520, 610, 480, 900, 870, 950)
t(sapply(split(p_ann, fine), function(v) c(median = median(v), IQR = IQR(v))))
```

Fine region 1 sits mostly (75 %) in coarse region 1; fine region 2 entirely in coarse 2.
The page's crosswalk lists every coarse region that holds at least 5 % of a fine region.

:::
